---
title: "exploratory"
format: html
editor: visual
---

This week’s goal was to clean all four raw datasets so they are ready for ingestion into a relational database in Week 3. The datasets include:

GRAPE_QUALITY.csv

vivino_wines_2026.csv

winemag-data_first150k.csv

winemag-data-130k-v2.csv

Each dataset required different cleaning steps depending on its structure and intended use. The cleaning focused on:

Ensuring consistent column names

Converting data types

Standardizing text fields

Removing duplicates

Handling missing values

Preparing each dataset for later normalization and schema design

```{r}
# Import Libraries

library(tidyverse)
library(lubridate)
library(janitor)
library(DBI)
library(duckdb)
```

```{r}
# Import Data

grape_quality <- read_csv("data/GRAPE_QUALITY.csv") %>% clean_names()
vivino <- read_csv("data/vivino_wines_2026.csv") %>% clean_names()
winemag_150k <- read_csv("data/winemag-data_first150k.csv") %>% clean_names()
winemag_130k <- read_csv("data/winemag-data-130k-v2.csv") %>% clean_names()

```

This dataset will eventually become my Observation table in my database, so it required the most cleaning

# Cleaning

## Cleaning Grape Quality Data

```{r}
# Grape_quality cleaning

grape_quality_clean <- grape_quality %>%
  mutate(
    harvest_date = parse_date_time(harvest_date, orders = c("ymd", "mdy", "dmy")),
    quality_category = str_to_title(quality_category),
    across(c(quality_score, sugar_content_brix, acidity_ph,
             cluster_weight_g, berry_size_mm, sun_exposure_hours),
           ~ as.numeric(.)
           )
    )

# standardizing text fields

grape_quality_clean <- grape_quality_clean %>%
  mutate(
    variety = variety |> str_squish() |> str_to_title(),
    region  = region  |> str_squish() |> str_to_title(),
    quality_category = quality_category |> str_squish() |> str_to_title()
  )

# Check for out of scope values

grape_quality_clean %>%
  summarise(
    min_brix = min(sugar_content_brix, na.rm = TRUE),
    max_brix = max(sugar_content_brix, na.rm = TRUE),
    min_ph   = min(acidity_ph, na.rm = TRUE),
    max_ph   = max(acidity_ph, na.rm = TRUE),
    min_sun  = min(sun_exposure_hours, na.rm = TRUE),
    max_sun  = max(sun_exposure_hours, na.rm = TRUE),
    min_moist = min(soil_moisture_percent, na.rm = TRUE),
    max_moist = max(soil_moisture_percent, na.rm = TRUE),
    min_rain = min(rainfall_mm, na.rm = TRUE),
    max_rain = max(rainfall_mm, na.rm = TRUE)
  )

```

## Cleaning Vivino Data

```{r}
# Vivino cleaning and ensure numeric fields and standardized text

vivino_clean <- vivino %>%
  mutate(
    rating = as.numeric(rating),
    ratings_count = as.integer(ratings_count),
    vintage_year = as.integer(vintage_year),
    across(
      c(full_name, wine_name, winery_name, wine_type, region, country),
      ~ .x |> str_squish() |> str_to_title()
    )
  ) %>%
  distinct()

```

## Clean Winemag Dataset

```{r}
# Add columns 150k doesnt have so we can concatenate both sets

winemag_150k <- winemag_150k %>%
  mutate(
    taster_name = NA_character_,
    taster_twitter_handle = NA_character_,
    title = NA_character_
  )

```

```{r}
#Wine mag clean function 

clean_winemag <- function(df) {
  df %>%
    clean_names() %>%
    mutate(
      price = as.numeric(price),
      points = as.integer(points),
      taster_twitter_handle = na_if(taster_twitter_handle, ""),
      country = str_to_title(country),
      province = str_to_title(province),
      region_1 = str_to_title(region_1),
      region_2 = str_to_title(region_2),
      variety = str_to_title(variety),
      winery = str_to_title(winery)
    )
}

```

```{r}
# Apply the changes and combine the datasets
winemag_130k_clean <- clean_winemag(winemag_130k)
winemag_150k_clean <- clean_winemag(winemag_150k)

# Bind them
winemag_all <- bind_rows(winemag_130k_clean, winemag_150k_clean)

```

```{r}
winemag_clean <- winemag_all %>%
  select(
    country, description, designation, points, price,
    province, region_1, region_2, taster_name,
    taster_twitter_handle, title, variety, winery
  ) %>%
  mutate(
    across(
      c(country, province, region_1, region_2, variety, winery, taster_name, title),
      ~ .x |> str_squish() |> str_to_title()
    )
  ) %>%
  distinct()

```

# Cleaning Summary

This week, I cleaned four wine‑related datasets to prepare them for ingestion into a relational database. Each dataset required different cleaning steps based on its structure and intended role in the final schema.

GRAPE_QUALITY.csv received the most extensive cleaning because it will become the core Observation table. I standardized text fields, converted numeric and date columns, checked for impossible values, and ensured consistent formatting across varieties and regions.

vivino_wines_2026.csv was cleaned to standardize winery names, wine types, and regions. Numeric fields were converted, and duplicates were removed. This dataset will later supply winery and wine‑type lookup tables.

winemag-data_first150k.csv and winemag-data-130k-v2.csv were aligned by adding missing columns, cleaned using a shared function, and merged into a single tidy dataset. Only relevant columns were kept, and text fields were standardized.

The above now enables us to ingest the data into the Duckdb for week 3.

# Exploratory

```{r}
unique(grape_quality$region)
```

```{r}
unique(vivino$region)
```

```{r}
unique(winemag_all$province)
```

```{r}
unique(grape_quality$variety)
```

This tells me what columns I would like to include in my database.

# Connect with the Duckdb

```{r}
con <- DBI::dbConnect(duckdb::duckdb(), "wine_database.duckdb")
```

# Export the Cleaned Dataframes to CSV

```{r}
dir.create("clean", showWarnings =  FALSE)
```

```{r}

write_csv(grape_quality_clean, "clean/grape_quality_clean.csv")
```

```{r}

vivino_clean <- vivino_clean %>%
  select(
    country,
    rating,
    wine_name,
    winery_name,
    ratings_count,
    wine_type,
    region,
    is_natural
  )

write_csv(vivino_clean, "clean/vivino_clean.csv")
```

```{r}
winemag_clean <- winemag_clean %>%
  select(
    variety,
    winery,
    country,
    province,
    region_1,
    points,
    price
  )

write_csv(winemag_clean, "clean/winemag_clean.csv")
```

# Create Tables

## Create Grape Quality Table

### In Case Table Drop Needed

```{sql connection=con}

-- in case we need to delete tables

DROP TABLE IF EXISTS grape_quality_clean;
DROP TABLE IF EXISTS vivino_clean;
DROP TABLE IF EXISTS winemag_clean;

```

```{sql connection=con}
-- Create grape quality table

CREATE TABLE grape_quality_clean (
  sample_id INTEGER,
  variety TEXT,
  region TEXT,
  quality_score DOUBLE,
  quality_category TEXT,
  sugar_content_brix DOUBLE,
  acidity_ph DOUBLE,
  cluster_weight_g DOUBLE,
  berry_size_mm DOUBLE,
  harvest_date DATE,
  sun_exposure_hours DOUBLE,
  soil_moisture_percent DOUBLE,
  rainfall_mm DOUBLE
);

```

## Create Vivino Table

```{sql connection=con}

CREATE TABLE vivino_clean (
  country TEXT,
  rating DOUBLE,
  wine_name TEXT,
  winery_name TEXT,
  ratings_count INTEGER,
  wine_type TEXT,
  region TEXT,
  is_natural BOOLEAN
);


```

## Create Winemag Table

```{sql connection=con}

CREATE TABLE winemag_clean (
  variety TEXT,
  winery TEXT,
  country TEXT,
  province TEXT,
  region_1 TEXT,
  points BIGINT,
  price DOUBLE
);


```

# Ingesting the Data

```{sql connection=con}
COPY grape_quality_clean
FROM 'clean/grape_quality_clean.csv'
DELIMITER ','
CSV HEADER;

```

```{sql connection=con}

COPY vivino_clean
FROM 'clean/vivino_clean.csv'
DELIMITER ','
CSV HEADER
-- NA values so well change the NA to NULL
NULL 'NA';

```

```{sql connection=con}

COPY winemag_clean
FROM 'clean/winemag_clean.csv'
DELIMITER ','
CSV HEADER
NULL 'NA';

```

# Verify Column types were correctly assigned

```{sql connection=con}

SELECT 'grape_quality_clean' AS table, COUNT(*) AS rows FROM grape_quality_clean
UNION ALL
SELECT 'vivino_clean', COUNT(*) FROM vivino_clean
UNION ALL
SELECT 'winemag_clean', COUNT(*) FROM winemag_clean;

```

```{sql connection=con}

DESCRIBE grape_quality_clean;

```

```{sql connection=con}

DESCRIBE vivino_clean;

```

```{sql connection=con}

DESCRIBE winemag_clean;

```

## Inspect Data

```{sql connection=con}

SELECT * FROM grape_quality_clean LIMIT 10;
```

```{sql connection=con}

SELECT * FROM vivino_clean LIMIT 10;

```

```{sql connection=con}

SELECT * FROM winemag_clean LIMIT 10;

```

# Exploratory Queries

## Grape Quality Queries

```{sql connection=con}

SELECT DISTINCT variety
FROM grape_quality_clean
ORDER BY variety;

```

```{sql connection=con}

SELECT region, AVG(quality_score) AS avg_quality
FROM grape_quality_clean
GROUP BY region
ORDER BY avg_quality DESC;

```

## Vivino Queries

```{sql connection=con}

SELECT wine_name, winery_name, rating
FROM vivino_clean
ORDER BY rating DESC
LIMIT 10;

```

## Wine Magazine Query

```{sql connection=con}

SELECT region_1, AVG(points) AS avg_points
FROM winemag_clean
GROUP BY region_1
ORDER BY avg_points DESC
LIMIT 10;

```

# Joining to Verify Relationships

```{sql connection=con}

SELECT
  g.variety,
  COUNT(*) AS grape_samples,
  COUNT(w.variety) AS winemag_reviews
FROM grape_quality_clean g
LEFT JOIN winemag_clean w
  ON g.variety = w.variety
GROUP BY g.variety
ORDER BY grape_samples DESC;
```

```{sql connection=con}

SELECT
  v.country,
  median(v.rating) AS avg_vivino_rating,
  median(w.points) AS avg_winemag_points
FROM vivino_clean v
JOIN winemag_clean w
  ON v.country = w.country
GROUP BY v.country
ORDER BY avg_vivino_rating DESC;
```

```{sql connection=con}

SELECT DISTINCT

```

# The Questions to Address

Questions to consider:

-   Which country has the highest Brix?
-   Which Which French Region/Province has the highest rating/Brix?
-   Which Region, grouped by variety have the highest brix?
-   Which variety, grouped by region, has the highest brix? Have a threshold of environmental factors?
